Developing decarbonisation pathways for Irish homes in changing TIMES

### 2021-08-26 manuscript.tex
26 Aug

\title{Developing Decarbonisation Pathways for Irish homes in changing TIMES}

@@

Ireland has committed to one of the most ambitious decarbonisation targets in the world to 2030. The residential sector, responsible for 14.2\% of total greenhouse gas (GHG) emissions \cite{Howley2020Energy-RelatedIreland.} , is among the most carbon intensive per dwelling in the EU, emitting 60\% more $CO_2$ than the national EU average \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, because of a high share of low thermal efficient detached dwellings coupled with a high dependency on carbon-intensive heating fuels. These factors have served as barriers to the decarbonisation of the residential sector in Ireland to date.  \par
Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges, which provides insights to decision-makers on issues related to climate and energy policy. \par
The TIMES-Ireland Model (TIM) is a newly developed optimisation model for the Irish energy system. This paper describes the methodology applied to the development of the residential sector in TIM and explores the results, which provides new insights on optimal residential decarbonisation pathways.

@@

The Introduction section of this paper reviews bottom-up residential models, which have high technological granularity accounting for the residential sector only and ESOMs which have a bottom-up or hybrid approach and includes all energy-related sectors. The hybrid approach combines bottom-up technological detail and top-down macroeconomic detail. The introduction also analyses Ireland's distinctive residential sector, climate targets, and current energy policy.

@@

Economically optimal or least cost residential decarbonisation pathways are dynamic, with technological advancements in heating appliances being assisted by exponentially increasing computational capacity over time, combined with new low-carbon fuel alternatives and new heating technologies becoming available. ESOMs have a rich representation of existing and future technological variables such as efficiency, cost, and fuel, while also accounting for macroeconomic variables such as economic growth and the social discount rate. ESOMs supply enough energy to at least satisfy energy service demands so that the lights stay on, while also considering sectoral interconnections to explore future decarbonisation pathways. There are a number of widely used ESOMs, such as MESSAGE \cite{Messner1995Model-basedPlanning}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.

These models can be used to explore the residential sector in higher granularity. Different levels of disaggregation are applied to the residential sector in models and research by \citet{Natarajana2011ModellingApproach}  highlighted a common approach used by defining an average performance for a number of dwelling categories, and this approach was adopted here. To account for the heterogeneous mix of dwellings, each dwelling is categorised based on building type to better reflect the housing stock and provide building specfic insights. Some models also disaggregate the year of construction \cite{Ahern2013StateStock}, wall type \cite{Dineen2015ImprovedSource} and level of insulation \cite{Clinch2001Cost-benefitEfficiency}.

@@

The Integrated MARKAL-EFOM System (TIMES) has been extensively used in both the UK and Ireland, and there are several examples of using TIMES to focus on the residential sector. TIMES assumes perfect foresight to enable a least-cost solution to be calculated, but \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. In light of this,  \citet{Li2018IncorporatingModel} investigated the heating technology preference in the UK using UK TIMES Model (UKTM) by incorporating a Discrete Choice Model (DCM) to better reflect how homeowners choose heating technologies and thus correcting for error in the perfect foresight assumption.

There is a rising tendency for ESOMs to focus on new building fabric \cite{Leibowicz2018OptimalServices,Dineen2011ModellingHeating} and retrofitting to thermally improve the existing residential stock \cite{Dineen2015ImprovedSource}. Another trend observed is the electrification of energy-related sectors \cite{Chiodi2013ModellingSystem}, while a focus on photovoltaic (PV) systems can be examined in parallel \cite{Marczinkowski2018ResidentialSystems}. Retrofitting and electrification are a fundamental part of residential decarbonisation in Ireland, and therefore play a central role in the development of the model described here. 
Furthermore, the new model includes technologies, which were not previously considered, one such technology is District Heating (DH), some initial studies have found it may be optimal to have 37\% of heat demand to be provided by DH in Ireland \cite{Thellufsen2019ImplementingIreland}, the full list of space and water heating technologies considered in the model can be found in Table \ref{AppHeatingTech}. The strength of an ESOM can be used to examine the sectoral interactions, one study by Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} examined the interaction between heat pumps and wind generation investments in Denmark, the results find it is optimal to install air-to-water heat pumps in non-DH areas.

@@

In 2014, \citet{Rogan2014LEAPsSystem} linked OSeMOYOS to the Long Range Energy Alternatives Planning system (LEAP), which is a scenario-based simulation modeling tool used to analysis energy policy in all sectors of the economy. The models were linked to explore specific energy efficiency policies in the residential and other sectors. Three scenarios were explored, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario. The reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020, and the results show that residential total energy consumption rose 6.7\% in 2020, compared to 2008. In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and low energy lighting, resulting in a 9.9\% reduction of energy consumption in 2020, compared to the reference scenario. The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits. This results in an energy consumption reduction of 16.5\% in 2020, compared to the reference scenario. \citet{Rogan2014LEAPsSystem} discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements. The insights highlighted the importance for the policy to not only look at the number of retrofits, but also the depth of retrofit, as shown by the different results in the scenarios. The model also found considerable scope for energy savings to be made through improved building regulations.

In 2018, \citet{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of DH in both the commercial and residential sectors of Ireland. The paper analyses the $CO_2-80$ scenario ( 80\% $CO_2$ reduction in 2050, compared to 1990 ) with and without DH using the previous Irish TIMES model and EnergyPLAN. The previous version of Irish TIMES did not account for DH, so EnergyPLAN is used to investigate the operation and  feasibility of DH in Ireland. A sensitivity analysis was performed on investment costs, heat grid losses, and discount rates. In all cases, the inclusion of DH was found to be favourable. The results find the conversion of 37\% of the Irish heat demand to DH to be optimal. Both the TIMES model and climate target used in this study are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of DH.

\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector models. These models have provided insights on the input data, methodology, and results, and some studies are outlined below.
\par In the first comprehensive analysis of the Irish residential sector, Clinch and Healy examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure}, occupancy behaviour \cite{Clinch2001ModellingEfficiency}, social benefit \& cost \cite{Clinch2001Cost-benefitEfficiency}  and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}.

\par
In 2011,  \citet{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP). The model focused on new buildings and examined the energy efficiency of space and water heating by examining technology and building fabric thermal efficiency.  %The paper acknowledges that retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperature effect on energy efficiency improvements. 
Later in 2015, \citet{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining the energy savings potential of energy efficiency improvement measures of the existing dwelling stock. The Dwelling Energy Assessment Procedure (DEAP) model was used, this is the national tool for Building Energy Rating (BER) calculation. The BER filters from this study were updated and applied in the model developed here. \citet{Dineen2015ImprovedSource} performed and sensitivity analysis and found that there was a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift. Concluding, BER standard assumptions are poor as they vastly overestimate dwelling energy consumption, the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.
\par In 2013, \citet{Ahern2013StateStock} investigated thermal retrofit measures of the existing detached, oil centrally heated, rural housing stock. %A base geometry according to age bands and thermal characteristics was created in the study. The default values used in some of the BER database variables where highlighted. For example, there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs despite a 2001 survey finding 18\% of detached housing in Ireland had no roof insulation. 
\citet{Ahern2013StateStock} found dwellings built post-2000 consume the greatest amount of energy (kWh/annum), due to their large floor area. Concluding that the majority of thermal energy efficiency progress in Irish dwellings is being offset. The model results show that thermal improvement measures resulted in a reduction of running costs and $CO_2$ emissions by an average of 65\% for houses constructed prior to 1979, compared to houses built in 1979 or after resulting in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures. This study highlights the low hanging fruit of pre-1979 oil centrally heated, rural housing stock. 
\par The Archetype Dwelling Energy Model - SQL (ArDEM-SQL) was developed by \citet{Uidhir2020ImprovingDecisions} ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and calculates the annual energy consumption for space/ water heating, and lighting using EN 13790:2004 to simulate the potential for improved energy efficiency gains within an existing retrofit program. Mac \citet{Uidhir2020ImprovingDecisions} examined the Better Energy Homes (BEH) retrofit grant scheme. 
%The five most common retrofit combinations were simulated for 9 distinct archetypes (3 BER energy rating groups, and 3 building types) and compared these to alternative retrofit combinations which delivered additional efficiency improvements. 
The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\%, relative to the set of retrofit combinations actually delivered from the program. This shows the the type of retrofit(s) and archetype can have a large affect on the energy saved.

\par Finally, the Sustainable Energy Authority of Ireland's (SEAI’s) National Energy Modelling Framework (NEMF) is a suite of modelling tools used for modelling the Irish energy system. The SEAI is Ireland's national energy authority and they play a key role in bridging the gap between grant/policy and consumer, by assisting consumers to reduce energy. The models listed are some of the models used within SEAI's NEMF to model the impact of incentives and opportunities created by specific policy instruments.

@@

The models described in this section include a range of national ESOMs and bottom-up residential models with varied objectives. The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switched to Microsoft SQL, with much higher computational power. The growing BER database is providing high data granularity input to Irish models. 
%The internal temperature is becoming more frequently analysed in Irish residential studies, a recurring theme is the wide range of internal temperatures which Dineen et al. \cite{Dineen2015ImprovedSource} found this assumption significantly changes the final residential energy consumption.

\subsection{Case study for Ireland}
\label{Intro:IreRSD}

@@

@@

\subsubsection{Policy context}
\label{Intro:Policy}
The Irish government's long-term overarching strategy aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}\cite{Ireland2018ProjectFramework}. The densification of Ireland's main cities will provide opportunities to improve the efficiency of heating, electricity, and transport within and between cities. 
\par The Marginal Abatement Cost Curve (MACC) produced in CAP2019, shows electricity \& transport are the most cost-effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}. In Ireland, the decarbonisation of electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2020Energy-RelatedIreland.}. 
Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private investment towards low carbon or renewable energy. A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}. Ireland's current carbon tax of €33.50/ $tCO_2$ is set to rise to €100/ $tCO_2$ in 2030. The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise renewable electricity investment to achieve legally binding renewable targets under 2009/28/EC. The Renewable Electricity Support Scheme (RESS) replaced REFIT in 2020. RESS supports renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target.

@@

The distinctive Irish residential sector provides an interesting case study and new Irish climate targets now means previous Irish models which relied on superseded climate targets are out-dated but not redundant, new up to date models are needed to explore the decarbonisation pathways and technical limits. No such ESOM with a focus on Irish residential sector exists, the strength of an ESOM is to account for sectoral interactions which bottom-up simulation models cannot provide.  Structural improvements and archetypes in the model aligns with existing retrofitting and electrification strategies. Other model improvements include filtering input data and calculating new energy service demands, while also considering internal temperatures. TIM will be a critical tool in exploring  
\textit{how changes to future technology developments affect the optimal residential heating decarbonisation pathway in Ireland? }

The layout of the rest of this paper is as follows; section \ref{LitR} is the core section of this paper and explains the methodology used to develop the residential sector. The treatment of input data in the model is described, the calculations are explained, and assumptions and constraints are highlighted. Section \ref{Results} outlines the scenarios used and displays results from the residential sector. Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models. Section \ref{Conclusion} draws some conclusions and discusses possible future work.

@@

\section{Methodology}
\label{LitR}

@@

\subsection{The TIMES-Ireland Model (TIM)}
\label{Overview}

@@

\subsection{Residential}
\label{LitR:RES}
The residential sector was developed in parallel with other sectors in TIM, and concurrently with the new Irish LEAP model, which allows for more accessible soft-linking to provide robust analytical insights by combining simulation and optimisation, and to exchange expertise between modelling teams. Figure \ref{fig:RES} illustrates the flow of energy from fuel to demand in the TIM residential sector.

@@

The BER database provides the base year bottom-up energy consumption data, this is scaled up to represent the entire stock and cross-checked with the top-down energy consumption from the 2018 SEAI Energy Balance. To compare from the two databases, fuels were were combined and condensed down to 13 fuels in TIM. Table \ref{AppTable:Fuels} shows an overview of the fuels combined from the different data sources.

\par In TIM, fuels such as ``coal", represent a combination of fuels from the SEAI Energy Balance and BER database. Gasoil which has similar properties to kerosene, is accounted for in the SEAI Energy Balance,  but there is no evidence to support keeping gasoil and kerosene separate would provide further insight into residential decarbonisation, therefore these fuels are combined in TIM. However, disaggregating ambient heat could potentially provide insights on residential decarbonisation, so the model has separated space ambient heat and water ambient heat. The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat, or wood. Through examining the SEAI Energy Balance and BER database, the fuel share for coal, peat, and wood in solid multi-fuel was set to 19.4\%, 74.4\% and 6.2\% respectively. 
\par There are two types of energy services represented in Fig. \ref{fig:RES} - Non-Archetype and Archetype. Essentially Non-archetype energy services have a lack of data and could not be accurately assigned to an archetype. The six non-archetype energy services are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service. Non-Archetype data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{LitR:Lighting,Appliances}. The four archetype energy services are dependent on  building type and data is sourced from the BER database, which is discussed in Section \ref{LitR:StockData}. The internal temperature adjustment for space heating is also detailed in Section \ref{LitR:StockData}. The demand for archetype and non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{LitR:Demands}.

\subsubsection{Dwelling Stock Data}
\label{LitR:StockData}

>>>>

\title{Developing decarbonisation pathways for Irish homes in changing TIMES}

@@

Ireland has committed to one of the most ambitious decarbonisation targets in the world to 2030. The residential sector, responsible for 14.2\% of total greenhouse gas (GHG) emissions \cite{Howley2020Energy-RelatedIreland.} , is among the most carbon intensive per dwelling in the EU, emitting 60\% more $CO_2$ than the national EU average \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, because of a high share of low thermal efficient detached dwellings coupled with a high dependency on carbon-intensive heating fuels. These factors have served as barriers to the decarbonisation of the residential sector in Ireland to date.  The TIMES-Ireland Model (TIM) is a newly developed optimisation model for the Irish energy system. This paper describes the innovative methodology applied to the development of the residential sector in TIM and explores the results, which provides new insights on optimal residential decarbonisation pathways.

@@

The Introduction section of this paper reviews Energy System Optimisation Models (ESOMs) which use a bottom-up or hybrid approach and include all energy-related sectors, and bottom-up residential sector models which have have a detailed level of technological representation. The hybrid approach combines bottom-up technological detail and top-down macroeconomic detail. The introduction also analyses Ireland's distinctive residential sector, climate targets, and current energy policy.

@@

Economically optimal or least cost residential decarbonisation pathways are dynamic, advancements in heating technologies, new low-carbon fuel alternatives and new heating technologies continuously alter the optimal pathway. ESOMs have a rich representation of existing and future technological variables such as efficiency, cost, and fuel, while also accounting for macroeconomic variables such as economic growth and the social discount rate. ESOMs supply enough energy to at least satisfy each energy service, while also considering sectoral interconnections to explore optimal decarbonisation pathways. There are a number of widely used ESOMs, such as MESSAGE \cite{Messner1995Model-basedPlanning}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.

These models can be used to explore the residential sector in high spatial, temporal and technological granularity. Different levels of disaggregation are applied to the residential sector in models, \citet{Natarajana2011ModellingApproach}  highlighted a common approach used by defining an average performance for a number of dwelling categories. To account for the heterogeneous mix of dwellings, each dwelling is categorised based on building type to better reflect the housing stock and provide building specific insights. Some models also disaggregate the year of construction \cite{Ahern2013StateStock}, wall type \cite{Dineen2015ImprovedSource} and level of insulation \cite{Clinch2001Cost-benefitEfficiency}.

@@

The Integrated MARKAL-EFOM System (TIMES) has been extensively used in both the UK and Ireland, and there are several examples of using TIMES to focus on the residential sector. TIMES assumes perfect foresight to enable a least-cost solution to be calculated, but \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. In light of this,  \citet{Li2018IncorporatingModel} investigated the heating technology preference in the UK using UK TIMES Model (UKTM) by incorporating a Discrete Choice Model (DCM) to better reflect how homeowners choose heating technologies and thus addressing for simplification of the perfect foresight assumption.

There is a rising tendency for ESOMs to focus on new building fabric \cite{Leibowicz2018OptimalServices,Dineen2011ModellingHeating} and retrofitting to thermally improve the existing residential stock \cite{Dineen2015ImprovedSource}. Another trend observed is the electrification of energy-related sectors \cite{Chiodi2013ModellingSystem}, while a focus on photovoltaic (PV) systems can be examined in parallel \cite{Marczinkowski2018ResidentialSystems}. The strength of an ESOM can be used to examine the sectoral interactions, one study by Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} examined the interaction between heat pumps and wind generation investments in Denmark, the results find it is optimal to install air-to-water heat pumps in non-DH areas.

@@

In 2014, \citet{Rogan2014LEAPsSystem} linked OSeMOYOS to the Long Range Energy Alternatives Planning system (LEAP), which is a scenario-based simulation modeling tool used to analysis energy policy in all sectors of the economy. The models were linked to explore specific energy efficiency policies in the residential and other sectors. Three scenarios were explored, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario, while three residential policies were analysed retrofitting, building regulation and lighting. At the time of this study little to no data on retrofits existed and the Building Energy Rating (BER) public database was much smaller. \citet{Rogan2014LEAPsSystem} research focuses on the theoretical energy savings from retrofits, \textit{with both the
number of dwellings and the scale of improvement
works specified exogenously from the model} \cite{Rogan2014LEAPsSystem}, the model limited retrofits to BER bands D to G while not accounting for internal temperature shifts.
The reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020, and the results show that residential total energy consumption rose 6.7\% in 2020, compared to 2008. In the energy efficiency scenario, 800,000 residential dwellings are retrofitted, resulting in a 9.9\% reduction of energy consumption compared to the reference scenario. The energy efficiency+ scenario allows for the same quantity of retrofits, but accounts for deeper retrofits resulting in an energy consumption reduction of 16.5\%, compared to the reference scenario. The insights highlight the importance of the depth of retrofit, as shown by the results. The model also found considerable scope for energy savings to be made through improved building regulations.

In 2018, \citet{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of District Heating (DH) in both the commercial and residential sectors of Ireland. The paper analyses the $CO_2-80$ scenario ( 80\% $CO_2$ reduction in 2050, compared to 1990 ) with and without DH using the previous Irish TIMES model and EnergyPLAN. The previous version of Irish TIMES did not account for DH, so EnergyPLAN is used to investigate the operation and  feasibility of DH in Ireland. A sensitivity analysis was performed on investment costs, heat grid losses, and discount rates. In all cases, the inclusion of DH was found to be favourable. The results find the conversion of 37\% of the Irish heat demand to DH to be optimal. Both the TIMES model and climate target used in this study are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of DH.

\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector models, which are not optimisation models. These models have provided insights on the input data, methodology, and results. In the first comprehensive analysis of the Irish residential sector, Clinch and Healy examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure}, occupancy behaviour \cite{Clinch2001ModellingEfficiency}, social benefit \& cost \cite{Clinch2001Cost-benefitEfficiency}  and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}.

\par
In 2011,  \citet{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP). The model focused on new buildings and examined the energy efficiency of space and water heating by examining technology and building fabric thermal efficiency.  %The paper acknowledges that retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperature effect on energy efficiency improvements. 
Later in 2015, \citet{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining the energy savings potential of energy efficiency improvement measures in the existing dwelling stock. The Dwelling Energy Assessment Procedure (DEAP) model was used, this is the national tool for the BER calculation. A complete set of filters was applied to the BER database before use in the model, erroneous and outlining data was removed. \citet{Dineen2015ImprovedSource} performed and sensitivity analysis and found that there was a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift. Concluding, BER standard assumptions are poor as they vastly overestimate dwelling energy consumption, the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.
\par In 2013, \citet{Ahern2013StateStock} investigated thermal retrofit measures of the existing detached, oil centrally heated, rural housing stock. %A base geometry according to age bands and thermal characteristics was created in the study. The default values used in some of the BER database variables where highlighted. For example, there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs despite a 2001 survey finding 18\% of detached housing in Ireland had no roof insulation. 
\citet{Ahern2013StateStock} found dwellings built post-2000 consume the greatest amount of energy (kWh/annum), due to their large floor area. Concluding that the majority of thermal energy efficiency progress in Irish dwellings is being offset. The model results show that thermal improvement measures resulted in a reduction of running costs and $CO_2$ emissions by an average of 65\% for houses constructed prior to 1979, compared to houses built in 1979 or after resulting in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures. This study highlights the potential energy savings within pre-1979 oil centrally heated, rural housing stock. 
\par The Archetype Dwelling Energy Model - SQL (ArDEM-SQL) developed by Mac \citet{Uidhir2020ImprovingDecisions} converts the Irish BER public database into Microsoft SQL language to calculate the annual energy consumption for space/ water heating, and lighting using EN 13790:2004 to simulate the potential for improved energy efficiency gains within an existing retrofit program. Mac \citet{Uidhir2020ImprovingDecisions} examined the Better Energy Homes (BEH) retrofit grant scheme. 
%The five most common retrofit combinations were simulated for 9 distinct archetypes (3 BER energy rating groups, and 3 building types) and compared these to alternative retrofit combinations which delivered additional efficiency improvements. 
The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\%, relative to the set of retrofit combinations actually delivered from the program. This shows the the type of retrofit(s) and archetype can have a large affect on the energy saved.

\par Finally, the Sustainable Energy Authority of Ireland's (SEAI’s) National Energy Modelling Framework (NEMF) is a suite of modelling tools used for modelling the Irish energy system. The SEAI is Ireland's national energy authority and they \textit{provide evidence-based advice across the government, state and semi-state bodies, in addition to delivering a range of grant and support programmes to promote sustainable energy} . The models listed are some of the models used within SEAI's NEMF to model the impact of incentives and opportunities created by specific policy instruments.

@@

The models described in this section include a range of national ESOMs and bottom-up residential models with varied objectives. The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switched to Microsoft SQL, the higher computational power allows for deeper analyses. The growing BER database is provides detailed open sourced data for input to Irish models. 
%The internal temperature is becoming more frequently analysed in Irish residential studies, a recurring theme is the wide range of internal temperatures which Dineen et al. \cite{Dineen2015ImprovedSource} found this assumption significantly changes the final residential energy consumption. 
\subsection{Case study for Ireland}
\label{Intro:IreRSD}

@@

\subsubsection{Policy context}
\label{Intro:Policy}

@@

The Irish government's long-term overarching strategy aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}\cite{Ireland2018ProjectFramework}. The densification of Ireland's main cities will provide opportunities to improve the efficiency of heating, electricity, and transport within and between cities. 
\par The Marginal Abatement Cost Curve (MACC) produced in CAP2019, shows electricity \& transport are the most cost-effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}. In Ireland, the decarbonisation of electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2020Energy-RelatedIreland.}. 
Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private investment towards low carbon or renewable energy. A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}. Ireland's current carbon tax of €33.50/ $tCO_2$ is set to rise to €100/ $tCO_2$ in 2030. The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise renewable electricity investment to achieve legally binding renewable targets under 2009/28/EC. The Renewable Electricity Support Scheme (RESS) replaced REFIT in 2020. RESS supports renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target.

@@

The distinctive Irish residential sector provides an interesting case study and new Irish climate targets now means previous Irish models which relied on superseded climate targets are out-dated, but not redundant. A new iteratively built transparent ESOM with wide stakeholder involvement, which is reflective of current climate policy is needed to robustly explore optimal decarbonisation pathways and technical limits. The new model will provide opportunity for deeper analysis of the energy system sectors, with accessible soft-linking capabilities. The model development process can be applied to any national or regional ESOM.  
No such ESOM with a focus on Irish residential sector exists, the strength of an ESOM is to account for sectoral interactions which bottom-up simulation models cannot provide. Structural improvements and archetypes in the model aligns with existing retrofitting and electrification strategies. Other model innovations include filtering the BER database, other EU countries with publicly available BER database under directive (EU) 2018/844 can apply similar methods, while TIM also considers internal temperatures shifts and new energy service demands were calculated. TIM will be a critical tool in exploring  
\textit{how changes to future technology developments affect the optimal residential heating decarbonisation pathway in Ireland? }

The layout of the rest of this paper is as follows; section \ref{Meth} is the core section of this paper and explains the methodology used to develop the residential sector. The treatment of input data in the model is described, the calculations are explained, and assumptions and constraints are highlighted. Section \ref{Results} outlines the scenarios used and displays results from the residential sector. Section \ref{Discussion} discusses the results, strengths, and weakness of the newly developed model and how it differs from other models. Section \ref{Conclusion} draws some conclusions and discusses possible future work.

@@

\section{Methodology}
\label{Meth}

@@

\subsection{TIMES}
\label{MethTIMES}

@@

\subsection{TIMES-Ireland Model (TIM)}
\label{Meth:TIM}

TIM was developed completely online during national lockdowns, with no face to face meetings, the model relied on git-centred model development process with different developers leading different sectors. Weekly meetings were combined with intensive daily ``sprint" meetings to keep model milestones on track. The transparency of TIM improved the iterative development process, external energy experts and other stakeholders contributed using web-based dashboards \footnote{https://tim-review1.netlify.app/results/}. The schematic in Figure \ref{fig:TIMOverviewRES} shows the main sectors of the model, represented in the demand-side modules and how they link with the supply-side modules, to create a full picture of the national energy system \textbf{SOURCE}.

\begin{figure}[!htbp]
 \centering
 \includegraphics[width=\textwidth]{Figures/TIMOverviewRES.jpg}
 \caption{Structure of TIM }
 \label{fig:TIMOverviewRES}
\end{figure}

\subsection{TIM: Residential Sector}
\label{Meth:TIMRes}

The residential sector was developed with awareness of the ongoing national retrofitting and electrification strategies and investments. The residential energy services were developed based upon regulation (EU) 431/2014, the BER database and EU databases. Figure \ref{fig:RES} illustrates the flow of energy from fuel to demand in the residential sector.

@@

The BER database provides the base year bottom-up energy consumption data, this is scaled up to represent the entire stock and cross-checked with the top-down energy consumption from the 2018 SEAI Energy Balance. To compare from the two databases, fuels were were combined and condensed down to 13 fuels in TIM. Table \ref{AppTable:Fuels} shows an overview of the fuels combined from the different data sources. The full list of space and water heating technologies considered in the model can be found in Table \ref{AppHeatingTech}.

\par In TIM, fuels such as ``coal", represent a combination of fuels from the SEAI Energy Balance and BER database. Gasoil which has similar properties to kerosene, is accounted for in the SEAI Energy Balance,  but there is no evidence to support keeping gasoil and kerosene separate would provide further insight into residential decarbonisation, therefore these fuels are combined in TIM. However, disaggregating ambient heat could potentially provide insights on residential decarbonisation, so the model has separated space ambient heat and water ambient heat. The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat, or wood. Through examining the SEAI Energy Balance and BER database, the fuel share for coal, peat, and wood in solid multi-fuel was set to 19.4\%, 74.4\% and 6.2\% respectively. 
\par There are two types of energy services represented in Fig. \ref{fig:RES} - Non-Archetype and Archetype. Essentially Non-archetype energy services have a lack of data and could not be accurately assigned to an archetype. The six non-archetype energy services are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service. Non-Archetype data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{Meth:Data}. The four archetype energy services are dependent on  building type and data is sourced from the BER database, which is discussed in Section \ref{Meth:Data}. The internal temperature adjustment for space heating is also detailed in Section \ref{Meth:Data}. The demand for archetype and non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{Meth:Data}.

\subsubsection{Data}
\label{Meth:Data}
%
The dwelling stock data is obtained from the SEAI BER Public database \footnote{https://ndber.seai.ie/BERResearchTool/ber/search.aspx}, which is firstly filtered, then scaled up using Central Statistics Office (CSO) data. 
The SEAI BER public database is Ireland's national Energy Performance Certificate (EPC) register,  which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD). The BERs are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). 
\par Over half of Irish dwellings are captured in the BER database and every BER displays over 200 energy related variables. The DEAP model which is used to assign BER labels,  is outlined by \citet{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM}. In terms of space heating, DEAP assumes that each dwelling is single zoned, the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season, which is from October to May inclusive and assumes monthly external temperatures, the average external temperature for the heating season is 7.6$^{\circ}$C. One assumption, the internal temperature assumption, results in a large error, where space and water heating are overestimated in DEAP. Space heating consumption is reduced by 26\% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance. DEAP assumes that living areas ( living room, sitting room, and possibly kitchen) in all dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C. Through the analysis of previous residential modelling in Ireland this is not the case, as the internal temperatures are often much lower, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions applied in TIM.
